Data Prep

### 2024-01-11 00.0_dataPrep.qmd
began changing variable names (through data cleaning step)

## Goals

Prepare data for the movement/social analysis

@@

-   `data/derived/inpadf_touse.feather`

```{r download-inpa}
#| eval: false
inpa <- move::getMovebankData(study = 6071688, login = MB.LoginObject, removeDuplicatedTimestamps = TRUE)
inpadf <- methods::as(inpa, "data.frame")
inpadf_touse <- inpadf %>%
  mutate(dateOnly = lubridate::ymd(substr(timestamp, 1, 10)),
         year = as.numeric(lubridate::year(timestamp))) %>%
  filter(lubridate::ymd(dateOnly) >= lubridate::ymd("2020-09-01"), lubridate::ymd(dateOnly) <= lubridate::ymd("2023-05-15")) # cut this off at the same point as the ornitela data
write_feather(inpadf_touse, "data/derived/inpadf_touse.feather")

length(unique(inpadf_touse$trackId)) # 77
```

@@

-   `data/derived/dat.feather`

```{r download-ornitela}
#| eval: false
# Ornitela: download data from movebank (just a subset of the times for now)
minDate <- "2020-09-01 00:00"
maxDate <- "2023-05-15 11:59"
dat <- vultureUtils::downloadVultures(loginObject = MB.LoginObject, removeDup = T, dfConvert = T, quiet = T, dateTimeStartUTC = minDate, dateTimeEndUTC = maxDate)
write_feather(ornitela, "data/derived/dat.feather")

length(unique(dat$trackId)) # 127
```

@@

-   `data/derived/inpadf_touse.feather`

-   `data/derived/dat.feather`

@@

-   `data/derived/dat3.feather`

```{r load-inpa-ornitela}
inpadf_touse <- read_feather("data/derived/inpadf_touse.feather")
dat <- read_feather("data/derived/dat.feather")
```

@@

# Will there be a problem joining these?
names(dat)[!(names(dat) %in% names(inpadf_touse))] # all names in "dat" are found in inpadf_touse
names(inpadf_touse)[!(names(inpadf_touse) %in% names(dat))] # many are not found in "dat" that are found in inpadf_touse.

# Add the dataset names so we can keep track of where the data comes from
inpadf_tojoin <- inpadf_touse[,names(dat)] %>%
  mutate(dataset = "inpa")
dat <- dat %>%
  mutate(dataset = "ornitela")

# Join the two datasets
joined <- bind_rows(inpadf_tojoin, dat)
dim(joined)

## fix trackId
dat2 <- joined %>%
  mutate(trackId = as.character(trackId),
         trackId = case_when(trackId == "E03" ~ "E03w",
                             TRUE ~ trackId))
# number of unique individuals
length(unique(dat2$trackId)) # 206

@@

# Are there any other individuals who have two different associated trackId's?
ww %>% filter(Movebank_id %in% dat2$trackId) %>% dplyr::select(Nili_id, Movebank_id) %>% distinct() %>% group_by(Nili_id) %>% summarize(n = length(unique(Movebank_id))) %>% arrange(desc(n)) # Note that there is one Nili_id in our dataset that has two associated trackId's--it's yomtov.
ww %>% filter(Nili_id =="yomtov") # Y26 and Y26b. So we should expect to once again "lose" an individual once we change to Nili ID's

all(dat2$trackId %in% ww$Movebank_id) # false.

# Okay, let's investigate the ones that aren't included
dat2 %>%
  filter(!(trackId %in% ww$Movebank_id)) %>%
  pull(trackId) %>% unique() # looks like this is going to be an issue of character matching. Let's transform everything to lowercase and remove all characters and spaces

dat2$trackId <- str_replace_all(tolower(dat2$trackId), "\\.", "")
ww$mb_id <- str_replace_all(tolower(ww$Movebank_id), "\\s|\\(|\\)|\\>", "") %>% str_replace_all(., "whiite", "white")

sort(unique(ww$mb_id))
unique(dat2$trackId)[!(unique(dat2$trackId) %in% ww$mb_id)]
dat2$trackId[dat2$trackId == "y01t60w"] <- "t60white"
dat2$trackId[dat2$trackId == "e86white"] <- "e86"
unique(dat2$trackId)[!(unique(dat2$trackId) %in% ww$mb_id)]
# Investigate the remaining ones:
dat2 %>% filter(!(trackId %in% ww$mb_id)) %>% dplyr::select(dateOnly, trackId, local_identifier, ring_id, dataset) %>% group_by(trackId, local_identifier, ring_id, dataset) %>% summarize(earlyDate = min(dateOnly)) # okay, looks like these three come from the INPA dataset and haven't been entered into the who's who yet.
# I guess for now let's just make their trackId's into Nili_id's

dat3 <- left_join(dat2, ww %>% dplyr::select(mb_id, Nili_id) %>% distinct(),
                  by = c("trackId" = "mb_id"))
nrow(dat3) == nrow(dat2) # good, same number of rows. (check that the join worked)
length(unique(dat3$Nili_id))
# For any that didn't get Nili_ids assigned, make it the trackId
dat3 %>% filter(is.na(Nili_id)) %>% pull(trackId) %>% unique() # as expecfted.
dat3 <- dat3 %>%
  mutate(Nili_id = case_when(is.na(Nili_id) ~ trackId,
                             TRUE ~ Nili_id))
unique(dat3$Nili_id)
write_feather(dat3, "data/derived/dat3.feather")
```

@@

-   `data/derived/dat3.feather`

@@

-   `datAnnot`

```{r load-hospital}
dat3 <- read_feather("data/derived/dat3.feather")
toRemove <- read_excel("data/raw/whoswho_vultures_20230315_new.xlsx", sheet = "periods_to_remove")
```

@@

## annotate the data with periods to remove
toRemove <- toRemove %>%
  dplyr::select(Nili_id,
                remove_start,
                remove_end,
                reason) %>%
  mutate(across(c(remove_start, remove_end), .fns = function(x){
    lubridate::ymd(x)
  })) %>%
  dplyr::filter(!is.na(remove_end))

toRemove_long <- toRemove %>%
  group_by(Nili_id, reason) %>%
  # sequence of daily dates for each corresponding start, end elements
  dplyr::mutate(dateOnly = map2(remove_start, remove_end, seq, by = "1 day")) %>%
  # unnest the list column
  unnest(cols = c(dateOnly)) %>%
  # remove any duplicate rows
  distinct() %>%
  dplyr::select(-c(remove_start, remove_end)) %>%
  rename("status" = reason)

# Join to the original data
datAnnot <- dat3 %>%
  left_join(toRemove_long, by = c("Nili_id", "dateOnly")) %>%
  mutate(status = replace_na(status, "valid"))
nrow(datAnnot) == nrow(dat3) #T: good, same number of rows.

# Actually REMOVE the periods to remove...
datAnnot <- datAnnot %>%
  filter(status == "valid")
nrow(datAnnot)
length(unique(datAnnot$Nili_id))
nrow(dat3)
nrow(datAnnot)
```

@@

@@

-   `datAnnot`

@@

```{r}
#| eval: false

# Identify and remove capture dates using Marta's code --------------------
## To identify the capture dates, first we need to classify the roosts (now using the get_roosts() function). Then, if the bird roosted within 500 m of the capture site, it was considered to be captured and that day, and the following day, are excluded from the dataset. This has been validated. Note: this protocol doesn't work for the Carmel because of the position of the roost sites relative to cages. But we're just dealing with southern individuals, so that's okay.
# roosts <- get_roosts_df(df = datAnnot, id = "Nili_id", timestamp = "timestamp", x = "location_long", y = "location_lat", ground_speed = "ground_speed", speed_units = "m/s", quiet = F)
# save(roosts, file = "data/derived/roosts.Rda")
load("data/derived/roosts.Rda")

@@

# Load the information about the capture sites
sites <- read.csv("data/raw/capture_sites.csv")

@@

## This, however, does not work for the Carmel, because the roosts are within 500m of the cage and very often the birds roost on top of the cage without actually being inside it. So for the Carmel captures, we will use another protocol: if the birds were sleeping within 50m of the cage and it was a capture day (or 3 days before the release day), we consider the birds were captured and we remove the data

@@

# It all looks ok, so we can subset the dataset to exclude the capture dates
datAnnot_noCaptures <- datAnnot %>%
  left_join(sub.captured.dates, by = c("Nili_id", "dateOnly" = "date"))
nrow(datAnnot) == nrow(datAnnot_noCaptures) # should be TRUE. NOW we can filter.
datAnnot_noCaptures <- datAnnot_noCaptures %>%
  dplyr::filter(Captured != "Yes"|is.na(Captured)) # remove the individual*days when they were captured

# How many rows and individuals did we remove?
before <- nrow(datAnnot)
after <- nrow(datAnnot_noCaptures)
(propChange <- (before-after)/before) # approx. 1% of data removed.
length(unique(datAnnot$Nili_id)) == length(unique(datAnnot_noCaptures$Nili_id)) # no change in number of individuals! As expected.

datAnnot_noCaptures <- datAnnot_noCaptures %>%
  dplyr::select(-c("ClosestCaptureSite", "ClosestCaptureDist", "Captured"))

write_feather(datAnnot_noCaptures, "data/derived/datAnnot_noCaptures.feather")
```

@@

-   `data/derived/datAnnot_noCaptures.feather`

@@

```{r}
datAnnot_noCaptures <- read_feather("data/derived/datAnnot_noCaptures.feather")
```

@@

datAnnot2 <- datAnnot_noCaptures %>%
  dplyr::select(-c("sex")) %>%
  left_join(as, by = "Nili_id")
nrow(datAnnot2) == nrow(datAnnot)
length(unique(datAnnot2$Nili_id)) # 196

datAnnot <- datAnnot2 # a hack so the rest of the code will work
```

@@

-   `datAnnot`

@@

-   `data/derived/datAnnotCleaned.Rda`

```{r}
#| eval: false
# Clean the data
## Region masking, downsampling, removal of speed outliers, setting altitude outliers to NA, etc.
mask <- sf::st_read("data/raw/CutOffRegion.kml")
datAnnotCleaned <- vultureUtils::cleanData(dataset = datAnnot, mask = mask, inMaskThreshold = 30, removeVars = F, idCol = "Nili_id", downsample = F, reMask = T) # XXX using 30 days overall. If we want 30 days per season, will have to do that differently (e.g. split the seasons earlier)--but I think that that'll be covered later in the data cleaning anyway.
length(unique(datAnnotCleaned$Nili_id)) # 172 individuals
# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
datAnnotCleaned <- datAnnotCleaned %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))

save(datAnnotCleaned, file = "data/derived/datAnnotCleaned.Rda")
load("data/derived/datAnnotCleaned.Rda")
```

@@

-   `datAnnotCleaned`

@@

```{r}
#| eval: false
# Split into Marta's 3 seasons --------------------------------------------
s1 <- datAnnotCleaned %>%
  mutate(start_breeding = ifelse(month %in% c(1:5),
                                 paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
                                 paste(as.character(year), "-", "12", "-", "15", sep = "")),
         start_summer = ifelse(month == 12,
                               paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
                               paste(as.character(year), "-", "05", "-", "15", sep = "")),
         start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
         start_breeding = as.Date(start_breeding),
         start_summer = as.Date(start_summer),
         season = case_when(
           dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
           dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
           dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
  filter(dateOnly != "2023-05-15") # remove May 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.

>>>>

## Goals and introduction

Prepare data for the movement/social analysis. This involves loading the data from Movebank and cleaning it.

@@

-   `data/derived/inpa.feather`

```{r download-inpa}
#| eval: false
inpa <- move::getMovebankData(study = 6071688, login = MB.LoginObject, removeDuplicatedTimestamps = TRUE)
inpa <- methods::as(inpa, "data.frame")
inpa <- inpa %>%
  mutate(dateOnly = lubridate::ymd(substr(timestamp, 1, 10)),
         year = as.numeric(lubridate::year(timestamp))) %>%
  filter(lubridate::ymd(dateOnly) >= lubridate::ymd("2020-09-01"), lubridate::ymd(dateOnly) <= lubridate::ymd("2023-05-15")) # cut this off at the same point as the ornitela data
write_feather(inpa, "data/derived/inpa.feather")

length(unique(inpa$trackId)) # 77
```

@@

-   `data/derived/ornitela.feather`

```{r download-ornitela}
#| eval: false
# Ornitela: download data from movebank (just a subset of the times for now)
minDate <- "2020-09-01 00:00"
maxDate <- "2023-05-15 11:59"
ornitela <- vultureUtils::downloadVultures(loginObject = MB.LoginObject, removeDup = T, dfConvert = T, quiet = T, dateTimeStartUTC = minDate, dateTimeEndUTC = maxDate)
write_feather(ornitela, "data/derived/ornitela.feather")

length(unique(ornitela$trackId)) # 127
```

@@

-   `data/derived/inpa.feather`

-   `data/derived/ornitela.feather`

@@

-   `data/derived/joined.feather`

```{r load-inpa-ornitela}
inpa <- read_feather("data/derived/inpa.feather")
ornitela <- read_feather("data/derived/ornitela.feather")
```

@@

# Will there be a problem joining these?
names(ornitela)[!(names(ornitela) %in% names(inpa))] # all names in "ornitela" are found in inpa
names(inpa)[!(names(inpa) %in% names(ornitela))] # many are not found in "ornitela" that are found in inpa

# Add the dataset names so we can keep track of where the data comes from
inpa_tojoin <- inpa[,names(ornitela)] %>%
  mutate(dataset = "inpa")
ornitela <- ornitela %>%
  mutate(dataset = "ornitela")

# Join the two datasets
joined0 <- bind_rows(inpa_tojoin, ornitela)
dim(joined0)

## fix trackId
joined0 <- joined0 %>%
  mutate(trackId = as.character(trackId),
         trackId = case_when(trackId == "E03" ~ "E03w",
                             TRUE ~ trackId))

# number of unique individuals
length(unique(joined0$trackId)) # 206

@@

# Are there any other individuals who have two different associated trackId's?
ww %>% filter(Movebank_id %in% joined$trackId) %>% dplyr::select(Nili_id, Movebank_id) %>% distinct() %>% group_by(Nili_id) %>% summarize(n = length(unique(Movebank_id))) %>% arrange(desc(n)) # Note that there is one Nili_id in our dataset that has two associated trackId's--it's yomtov.
ww %>% filter(Nili_id =="yomtov") # Y26 and Y26b. So we should expect to once again "lose" an individual once we change to Nili ID's

all(joined0$trackId %in% ww$Movebank_id) # false.

# Okay, let's investigate the ones that aren't included
joined0 %>%
  filter(!(trackId %in% ww$Movebank_id)) %>%
  pull(trackId) %>% unique() # looks like this is going to be an issue of character matching. Let's transform everything to lowercase and remove all characters and spaces

joined0$trackId <- str_replace_all(tolower(joined0$trackId), "\\.", "")
ww$mb_id <- str_replace_all(tolower(ww$Movebank_id), "\\s|\\(|\\)|\\>", "") %>% str_replace_all(., "whiite", "white")

sort(unique(ww$mb_id))
unique(joined0$trackId)[!(unique(joined0$trackId) %in% ww$mb_id)]
joined0$trackId[joined0$trackId == "y01t60w"] <- "t60white"
joined0$trackId[joined0$trackId == "e86white"] <- "e86"
unique(joined0$trackId)[!(unique(joined0$trackId) %in% ww$mb_id)]
# Investigate the remaining ones:
joined0 %>% filter(!(trackId %in% ww$mb_id)) %>% 
  dplyr::select(dateOnly, trackId, local_identifier, ring_id, dataset) %>%
  group_by(trackId, local_identifier, ring_id, dataset) %>% 
  summarize(earlyDate = min(dateOnly)) # okay, looks like these three come from the INPA dataset and haven't been entered into the who's who yet.
# I guess for now let's just make their trackId's into Nili_id's

joined <- left_join(joined0, ww %>% dplyr::select(mb_id, Nili_id) %>% distinct(),
                  by = c("trackId" = "mb_id"))
nrow(joined) == nrow(joined0) # good, same number of rows. (check that the join worked)
length(unique(joined$Nili_id))
# For any that didn't get Nili_ids assigned, make it the trackId
joined %>% filter(is.na(Nili_id)) %>% pull(trackId) %>% unique() # as expecfted.
joined <- joined %>%
  mutate(Nili_id = case_when(is.na(Nili_id) ~ trackId,
                             TRUE ~ Nili_id))
unique(joined$Nili_id)
write_feather(joined, "data/derived/joined.feather")
```

@@

-   `data/derived/joined.feather`

@@

-   `removed`

```{r load-hospital}
joined <- read_feather("data/derived/joined.feather")
toRemove <- read_excel("data/raw/whoswho_vultures_20230315_new.xlsx", sheet = "periods_to_remove")
```

@@

## annotate the data with periods to remove
toRemove <- toRemove %>%
  dplyr::select(Nili_id,
                remove_start,
                remove_end,
                reason) %>%
  mutate(across(c(remove_start, remove_end), .fns = function(x){
    lubridate::ymd(x)})) %>%
  dplyr::filter(!is.na(remove_end)) %>%
  group_by(Nili_id, reason) %>%
  # sequence of daily dates for each corresponding start, end elements
  dplyr::mutate(dateOnly = map2(remove_start, remove_end, 
                                seq, by = "1 day")) %>%
  # unnest the list column
  unnest(cols = c(dateOnly)) %>%
  # remove any duplicate rows
  distinct() %>%
  dplyr::select(-c(remove_start, remove_end)) %>%
  rename("status" = reason)

# Join to the original data
removal_annotated <- joined %>%
  left_join(toRemove, by = c("Nili_id", "dateOnly")) %>%
  mutate(status = replace_na(status, "valid"))
nrow(removal_annotated) == nrow(joined) #T: good, same number of rows.

# Actually REMOVE the periods to remove...
removed <- removal_annotated %>%
  filter(status == "valid")
nrow(removed)
length(unique(removal_annotated$Nili_id))
nrow(joined)
nrow(removal_annotated)
```

@@

To identify the capture dates, first we need to classify the roosts (now using the get_roosts() function). Then, if the bird roosted within 500m of the capture site, it was considered to be captured and that day, and the following day, are excluded from the dataset. This has been validated. This, however, does not work for the Carmel, because the roosts are within 500m of the cage and very often the birds roost on top of the cage without actually being inside it. So for the Carmel captures, we will use another protocol: if the birds were sleeping within 50m of the cage and it was a capture day (or 3 days before the release day), we consider the birds were captured and we remove the data

@@

-   `removed`

-   `data/raw/capture_sites.csv`

@@

```{r remove-capture-cage}
#| eval: false

# Identify and remove capture dates using Marta's code --------------------
roosts <- get_roosts_df(df = removed, id = "Nili_id", timestamp = "timestamp", x = "location_long", y = "location_lat", ground_speed = "ground_speed", speed_units = "m/s", quiet = F)
save(roosts, file = "data/derived/roosts.Rda")
load("data/derived/roosts.Rda")

@@

# Load information about the capture sites
sites <- read.csv("data/raw/capture_sites.csv")

@@

## For Carmel--different protocol.

@@

# It all looks ok, so we can subset the dataset to exclude the capture dates
removed_captures <- removed %>%
  left_join(sub.captured.dates, by = c("Nili_id", "dateOnly" = "date"))
nrow(removed) == nrow(removed_captures) # should be TRUE. NOW we can filter.
removed_captures <- removed_captures %>%
  dplyr::filter(Captured != "Yes"|is.na(Captured)) # remove the individual*days when they were captured

# How many rows and individuals did we remove?
before <- nrow(removed)
after <- nrow(removed_captures)
(propChange <- (before-after)/before) # approx. 1% of data removed.
length(unique(removed$Nili_id)) == length(unique(removed_captures$Nili_id)) # no change in number of individuals! As expected.

removed_captures <- removed_captures %>%
  dplyr::select(-c("ClosestCaptureSite", "ClosestCaptureDist", "Captured"))

write_feather(removed_captures, "data/derived/removed_captures.feather")
```

@@

-   `data/derived/removed_captures.feather`

@@

```{r attach-age-sex}
removed_captures <- read_feather("data/derived/removed_captures.feather")
```

@@

# Whenever I do left_joins, I save the result as something different before reassigning it to the variable name so I can make sure the join worked before committing to it. Hence the temporary "test" object.
test <- removed_captures %>%
  dplyr::select(-c("sex")) %>%
  left_join(as, by = "Nili_id")
nrow(test) == nrow(removed_captures)
length(unique(removed_captures$Nili_id)) # 196
removed_captures <- test
rm(test) # don't need this anymore after we know the join worked.
```

@@

-   `removed_captures`

@@

-   `data/derived/cleaned.Rda`

```{r clean-data}
#| eval: false
# Clean the data
## Region masking, downsampling, removal of speed outliers, setting altitude outliers to NA, etc.
mask <- sf::st_read("data/raw/CutOffRegion.kml")
cleaned <- vultureUtils::cleanData(dataset = removed_captures, mask = mask, inMaskThreshold = 30, removeVars = F, idCol = "Nili_id", downsample = F, reMask = T) # XXX using 30 days overall. If we want 30 days per season, will have to do that differently (e.g. split the seasons earlier)--but I think that that'll be covered later in the data cleaning anyway.
length(unique(cleaned$Nili_id)) # 172 individuals
# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
cleaned <- cleaned %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))

save(cleaned, file = "data/derived/cleaned.Rda")
load("data/derived/cleaned.Rda")
```

@@

-   `cleaned`

@@

```{r split-seasons}
#| eval: false
# Split into Marta's 3 seasons --------------------------------------------
s1 <- datAnnotCleaned %>%
  mutate(start_breeding = ifelse(month %in% c(1:5),
                                 paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
                                 paste(as.character(year), "-", "12", "-", "15", sep = "")),
         start_summer = ifelse(month == 12,
                               paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
                               paste(as.character(year), "-", "05", "-", "15", sep = "")),
         start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
         start_breeding = as.Date(start_breeding),
         start_summer = as.Date(start_summer),
         season = case_when(
           dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
           dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
           dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
  filter(dateOnly != "2023-05-15") # remove May 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.
